---
title: "Clonality in Prostanthera gilesii"
author: "Rose Andrew"
date: "28 August 2019"
output:
  html_document: default
editor_options:
  chunk_output_type: inline
---

```{r setup,echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, include=FALSE)
  
##################################################################
##################################################################
## Exploring clonality in DArTseq data
##################################################################
##################################################################
## Requires *SNP_2.csv file supplied by DArT and matching sample
## *metadata.csv in a folder specific to the DArT run (within 
## the data directory). 
##
##################################################################
## Load packages
##################################################################

library(hierfstat)
library(StAMPP)
library(ggplot2)
library(dartR)
library(poppr)
library(ape)
library(phangorn)

##################################################################
## Setup
##################################################################
## Specify main parameters
setwd(".")
overwrite=FALSE
## Directories: choose an indir and an analysis name
# indir="data/OrderAppendix_1_DPro19-4334" ## Psubset
indir="data/OrderAppendix_2_DPro19-4334" ## PgOnly
# indir="data/OrderAppendix_3_DPro19-4334" ## PgRels
# analysis_name="Psubset"
analysis_name="Pg"
# analysis_name="PgRels"

## Calculate other parameters
dartname=list.files(indir,"SNP_2.csv")
dartfile=file.path(indir,dartname)
metadataname=list.files(indir,"metadata.csv")
metadatafile=file.path(indir,metadataname)
outdir=file.path("output",analysis_name)
if(!dir.exists("output")) dir.create("output")
if(!dir.exists("temp")) dir.create("temp")
if(dir.exists(outdir)){
  if(overwrite){
    print("Overwriting existing directory")
  }else print("Directory already exists")
} else dir.create(outdir)



```
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,include=FALSE,results="hide")
                      
## Clone identification

The extent of clonality in Prostanthera gilesii was exlored using a range of genetic distance thresholds. This approach was necessary, as over 10,000 high-quality SNPs were polymorphic within the dataset. 

The first step was to examine the distribution of pairwise genetic distance (pairwise SNP divergence). Not surprisingly, genetic distances were highly bimodal across samples identified a priori as P. gilesii. Because it is likely that the sample from The Walls differs from the main Towac Ck population, it was excluded from further within-P. gilesii analyses.


```{r load, echo=FALSE, include=TRUE}

##################################################################
## Load data and convert to snpclone object
##################################################################
## Load the filtered data file, a genlight object called mygl1,
##  saved as a .rda file.
load(file = file.path("temp",paste0(analysis_name,"_filtered.rda")))

hist(mlg.filter(as.snpclone(mygl1),threshold = 1e-6,stats = "DISTANCES"),breaks = 100)

```


After excluding the sample from The Walls and filtering out invariant loci, all genetic distances were  below 11%.

Selecting a threshold of genetic distance above which genotypes are treated as distinct is somewhat arbitrary, as mutations accumulate continuously over time. The goal should be to distinguish samples that differ due to recombination of genotypes during sexual reproduction from those that differ only due to mutation. That is relatively simple in genetically diverse, outcrossing, as the genetic relatedness between even full siblings and between parents and offspring is half that between clones. If genetic diversity is low, as is the case in P. gilesii, many loci will be homozygous, so a larger number of loci is needed for reliable estimates of genetic distance. DArTseq provides this: after removing the sample from The Walls, the number of polymorphic loci is reduced by ~80%, but 2126  should provide a reasonable amount of power.


```{r filter, echo=FALSE, include=FALSE}

## Exclude the sample from the Walls
mygl2=popsub(mygl1,blacklist = "Walls")
nLoc(mygl2);nInd(mygl2)
mygl2=gl.filter.repavg(mygl2,threshold = 0.95)
nLoc(mygl2);nInd(mygl2)
mygl2=gl.filter.callrate(mygl2,method = "loc",threshold = 0.95)
mygl2=gl.filter.callrate(mygl2,method = "ind",threshold = 0.9)
nLoc(mygl2);nInd(mygl2)
mygl2=gl.filter.secondaries(mygl2)
nLoc(mygl2);nInd(mygl2)
mygl2=gl.filter.monomorphs(mygl2)
nLoc(mygl2);nInd(mygl2)

mygl2=gl.recalc.metrics(mygl2)
save(mygl2,file = file.path("temp",paste0(analysis_name,"_filtered_noWalls.rda")))

## Convert to snpclone object
mysc <- as.snpclone(mygl2)


```


```{r , echo=FALSE, include=TRUE}
##################################################################
## Find natural genetic distance thresholds
##################################################################
## First plot genetic distance histogram, and effects of different
##  thresholds and algorithms - needed to tweak input to get decent plot
hist(mlg.filter(mysc,threshold = 1e-6,stats = "DISTANCES"),breaks = 100)
```

We can see the effect of different distance thresholds on genotype identification using different approaches to clustering.

```{r ,  echo=FALSE, include=TRUE}
plot_filter_stats(mysc,filter_stats(mysc,threshold = .1),
                  mlg.filter(mysc,threshold=0,stats = "DISTANCES"),
                  breaks = "scott")

```



```{r ,  echo=FALSE, include=FALSE}

## Then see if the automated approach gives a similar result
cutoff_predictor(filter_stats(mysc)$farthest$THRESHOLDS)

## Cutoffs to use: 
auto_threshold=0.013
empir1_threshold=0.07
empir2_threshold=0.045
empir3_threshold=0.03
empir4_threshold=0.022

```

## Applying thresholds to the data

Based on the histogram above, we'll use a few different thresholds and compare the number of distinct genotypes identified. The second of the empirical thresholds seems to distinguish the major clusters conservatively.

```{r}

##################################################################
## Apply thresholds to data 
##################################################################
auto_mlg <- mlg.filter(mysc,threshold = auto_threshold)
auto_mysc <- mysc
auto_mysc$mlg <- auto_mlg
empir1_mlg <- mlg.filter(mysc,threshold = empir1_threshold)
empir1_mysc <- mysc
empir1_mysc$mlg <- empir1_mlg
empir2_mlg <- mlg.filter(mysc,threshold = empir2_threshold)
empir2_mysc <- mysc
empir2_mysc$mlg <- empir2_mlg
empir3_mlg <- mlg.filter(mysc,threshold = empir3_threshold)
empir3_mysc <- mysc
empir3_mysc$mlg <- empir3_mlg
empir4_mlg <- mlg.filter(mysc,threshold = empir4_threshold)
empir4_mysc <- mysc
empir4_mysc$mlg <- empir4_mlg


sclist <- list(empir1_mysc,empir2_mysc,empir3_mysc,empir4_mysc,auto_mysc)
mlglist <- sapply(sclist,mlg)

mytab=data.frame(Threshold=c("empirical 1","empirical 2","empirical 3","empirical 4","auto"),Cutoff=c(empir1_threshold,empir2_threshold,empir3_threshold,empir4_threshold,auto_threshold), 
                 nGenotypes=mlglist
                 )

```


```{r , echo=FALSE, include=TRUE}
print(mytab)
```

## Explore clonal lineages

By plotting the genotypes identified with different thresholds on the the PCoA plot, we can see which thresholds make sense. The second of the empirical thresholds seems to distinguish the major clusters conservatively.
Lots of plots!

```{r , echo=FALSE, include=TRUE, results="hide"}
## plot on PCoA
mypc <- gl.pcoa(mygl2,nfactors = 6)
## Scree plot
barplot(mypc$eig/sum(mypc$eig)*100)

mygl2@pop=as.factor(empir1_mlg)
gl.pcoa.plot(mypc,mygl2,labels="pop",xaxis=1,yaxis=2)+ggtitle("Empirical threshold 1")
mygl2@pop=as.factor(empir2_mlg)
gl.pcoa.plot(mypc,mygl2,labels="pop",xaxis=1,yaxis=2)+ggtitle("Empirical threshold 2")
gl.pcoa.plot(mypc,mygl2,labels="pop",xaxis=1,yaxis=3)+ggtitle("Empirical threshold 2")
gl.pcoa.plot(mypc,mygl2,labels="pop",xaxis=1,yaxis=4)+ggtitle("Empirical threshold 2")
gl.pcoa.plot(mypc,mygl2,labels="pop",xaxis=1,yaxis=5)+ggtitle("Empirical threshold 2")
gl.pcoa.plot(mypc,mygl2,labels="pop",xaxis=1,yaxis=6)+ggtitle("Empirical threshold 2")
mygl2@pop=as.factor(empir3_mlg)
gl.pcoa.plot(mypc,mygl2,labels="pop",xaxis=1,yaxis=2)+ggtitle("Empirical threshold 3")
mygl2@pop=as.factor(empir4_mlg)
gl.pcoa.plot(mypc,mygl2,labels="pop",xaxis=1,yaxis=2)+ggtitle("Empirical threshold 4")
mygl2@pop=as.factor(auto_mlg)
gl.pcoa.plot(mypc,mygl2,labels="pop",xaxis=1,yaxis=2)+ggtitle("Automatic threshold")

```

We can also use networks to examine the effects of decreasing the threshold.

```{r msn, echo=FALSE, include=TRUE, results="hide"}
## Plot phylo
plot.phylo(upgma(mlg.filter(mysc,threshold=0,stats = "DISTANCES")))

## Minimum spanning networks, with various thresholds
poppr.msn(empir1_mysc,mlg.filter(mysc,threshold=0,stats = "DISTANCES"))
poppr.msn(empir2_mysc,mlg.filter(mysc,threshold=0,stats = "DISTANCES"))
poppr.msn(empir3_mysc,mlg.filter(mysc,threshold=0,stats = "DISTANCES"))
poppr.msn(empir4_mysc,mlg.filter(mysc,threshold=0,stats = "DISTANCES"))
poppr.msn(auto_mysc,mlg.filter(mysc,threshold=0,stats = "DISTANCES"))

```



## Clonal diversity statistics
We can now examine the clonal diversity within each of the sample sets. Although few samples were from outside of the main Towac Ck population, it is instructive to include them, as the diversity of the ex-situ collections is an important question.


```{r distribution,include=TRUE,echo=FALSE}

## Tabulate data
mytable=mlg.table(empir2_mysc)



# print(as.data.frame(mytable))

```

We can also calculate diversity statistics, with confidence intervals estimated using bootstrap resampling. The statistics offered are H=Shannon's Index, G=Stoddart and Taylor's Index, lambda=Simpson's index, E5=Evenness, and CF=clonal fraction (the proportion of pairs that are clones). 


```{r diversity, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}
## Function for clonal fraction (from https://cran.r-project.org/web/packages/poppr/vignettes/mlg.html)
myCF <- function(x){
  x <- drop(as.matrix(x))
  if (length(dim(x)) > 1){ # if it's a matrix
    res <- rowSums(x > 0)/rowSums(x)
  } else {                 # if it's a vector
    res <- sum(x > 0)/sum(x)
  }
  return(res)
}

## Table with clonal fraction added
# diversity_stats(mytable, CF = myCF)

mytableCI <- diversity_ci(mytable, n = 100L, rarefy = FALSE,plot=T, raw = FALSE, CF = myCF) ## Not sure why the plot isn't printing

# print(as.data.frame(mytableCI)) 
```

Standard population summary statistics can give us insight into the relationships of individuals. 
```{r sumstats, echo=FALSE, include=TRUE}

## Summary statistics
sumstats=basic.stats(cbind(empir2_mysc@pop,as.data.frame(empir2_mysc)))
# sumstats$overall
pop_summary=data.frame(N=colMeans(sumstats$n.ind.samp,na.rm=T),
                       Ho=colMeans(sumstats$Ho,na.rm=T),
                       He=colMeans(sumstats$Hs,na.rm=T),
                       Fis=colMeans(sumstats$Fis,na.rm=T))
pop_summary


```

The observed heterozygosity (Ho) is higher than that expected at Hardy-Weinberg equilibrium, leading to negative inbreeding coefficients. This is consistent with high clonality, as large clones can share identical heterozygous genotypes, which are unlikely in outbreeding populations. Fixed heterozygosity is a feature of clones, and a high proportion of loci with heterozygosity approaching 1. However, we do not know the ploidy level of this species, and similar patterns of diversity could be produced by homeologs (or paralogs).

```{r ho_hist, echo=FALSE, include=TRUE}
hist(sumstats$perloc$Ho)
plot(sumstats$perloc$Hs,sumstats$perloc$Ho)

```


